# Rage Generation {#rage}

```{r setup-rage, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
theme_set(theme_minimal())
set.seed(1234)
clusters <- read_csv("data/derived/swing_clusters.csv")
swings   <- read_csv("data/derived/swings.csv")
energize <- read_csv("data/derived/energize.csv")
procs    <- read_csv("data/derived/procs.csv")
ubw5     <- filter(procs, ubw_rank == 5)
ubw_rate <- sum(ubw5$ubw_procs) / sum(ubw5$landed_white)   # Unbridled Wrath proc rate at 5/5
meta     <- yaml::read_yaml("data/logs/metadata.yaml")

conv <- function(lvl) 0.0091107836 * lvl^2 + 3.225598133 * lvl + 4.2652911

# Speed and hand for each swing: the posted weapon speed nearest to the measured swing interval,
# and main vs off hand from the rate (off-hand rates are about half the main-hand ones).
weapons <- map_dfr(meta, ~ tibble(file = .x$file, speed = map_dbl(.x$equipped_weapons, "speed"),
                                  weapon = map_chr(.x$equipped_weapons, "name")))
cl <- clusters %>%
  mutate(speed = map2_dbl(file, swing_interval_median,
                          ~ { s <- weapons$speed[weapons$file == .x]; s[which.min(abs(s - .y))] }),
         rate  = mean_gain / speed,
         hand  = case_when(setup == "two_hand" ~ "Two-hand", rate > 2.5 ~ "Main hand", TRUE ~ "Off hand"))

obs <- swings %>% filter(!capped) %>%
  mutate(gain = gain / 10) %>%
  left_join(cl %>% select(file, lo = mean_gain, speed, hand), by = "file", relationship = "many-to-many") %>%
  filter(abs(gain - lo) <= 0.2) %>%
  left_join(weapons, by = c("file", "speed")) %>%
  mutate(classic = 15 * damage / conv(level),
         group   = paste0(character, ", ", hand, " ", speed))

fmt_p <- function(p) format.pval(p, digits = 2, eps = 1e-10)

# Chapter version: newest first. See "Changelog" at the end of the chapter for what each part means.
changelog <- tribble(
  ~version, ~date,        ~change,
  "2.1.0",  "2026-09-23", "Added a log from Tarch with Unbridled Wrath at 4/5 (dual wield and two-handed), which completes the proc rate at every talent rank. 4/5 procs on about 36% of hits, still well below the tooltip's 48%.",
  "2.0.0",  "2026-09-23", "Added six logs from Tarch (Unbridled Wrath at 1/5, 2/5 and 3/5, dual wield and two-handed). Unbridled Wrath gives about 8% per talent point instead of the tooltip's 12%, so 5/5 is about 36% instead of 60% (a bug, not a talent-point mix-up). A second two-handed weapon (3.3 speed) confirms 4.5 × speed.",
  "1.1.0",  "2026-09-23", "Measured how often Unbridled Wrath procs (about 36% of landed white hits, well below the intended 60%).",
  "1.0.0",  "2026-09-23", "First version: three logs from beta build 1.60.1.69977, with p-values, a Bayesian estimate of the rates, and a level 60 simulation."
)
```

*Version `r changelog$version[1]` (`r changelog$date[1]`), based on `r length(meta)` logs from beta build 1.60.1. See the [changelog](#rage-changelog) for what changed.*

## Base hypothesis: the Classic formula

A white hit is a normal auto attack (so not an ability like Heroic Strike). In Classic, it's pretty well documented that the rage from a white hit depended on the damage of that hit. The Classic compendium used that formula:

\begin{equation}
Rage = \frac{D}{C} \times 15
\end{equation}

Where:

- $Rage$ is the rage that one white hit gives.
- $D$ is the damage of that hit.
- $C$ is the rage conversion value, basically a constant that increases with your level. At level 60 it is about 230, and at level 15 it is about 55.
- 15 is a fixed number from the game.

The game calculates $C$ from your level, $Lvl$:

\begin{equation}
C = 0.0091107836 \times Lvl^2 + 3.225598133 \times Lvl + 4.2652911
\end{equation}

Example: You are a level 15 warrior and hit a mob for 110 damage. The formula gives $110 / 55 \times 15 = 30$ rage.

This formula makes three predictions for World of Warcraft: Forever:

1. A hit that does more damage gives more rage.
2. A critical hit does about twice the damage, so it gives about twice the rage.
3. At a higher level, the same damage gives less rage, because $C$ is larger.

Moving forward, we'll use these predictions as our starting hypothesis.

## The logs

So far we've looked at `r length(meta)` combat logs, some posted to the [forever-warrior discussions](https://github.com/magey/forever-warrior/discussions) and some shared with us directly, all on beta build 1.60.1. The nice thing about these logs is that they record the warrior's current rage on almost every event, so we can read the rage from a single white swing straight off the log instead of having to guess at it.

```{r}
tibble(
  Log       = map_chr(meta, "character_name"),
  Setup     = map_chr(meta, "setup"),
  Weapons   = map_chr(meta, ~ paste0(map_chr(.x$equipped_weapons, "name"), " (",
                                   map_dbl(.x$equipped_weapons, "speed"), ")", collapse = " / ")),
  Source    = map_chr(meta, "source")
) %>% knitr::kable(caption = "Logs and the metadata posted with them (main hand first).")
```

We left out any swing made while sitting at 100 rage, since the cap cuts off whatever that swing would have given.

## Findings

```{r}
clusters %>%
  transmute(Character = character, Setup = setup,
            `Swing interval (s)` = round(swing_interval_median, 2),
            `Rage per swing` = round(mean_gain, 2),
            `Rage per second of swing interval` = round(mean_gain / swing_interval_median, 2),
            `Damage range` = paste(damage_min, damage_max, sep = "-"),
            `Crit` = round(mean_gain_crit, 2), `Non-crit` = round(mean_gain_noncrit, 2),
            Swings = n) %>%
  knitr::kable(caption = "Rage per white swing, grouped by the distinct values observed.")
```

```{r fig.cap="Rage per swing against damage dealt. Each group is flat: damage does not change the rage gained."}
swings %>% filter(!capped) %>%
  ggplot(aes(damage, gain / 10, colour = character)) +
  geom_jitter(height = 0.05, alpha = 0.5) +
  facet_wrap(~ setup) +
  labs(x = "Damage dealt", y = "Rage gained", colour = NULL)
```

1. **The Classic hypothesis doesn't hold.** How much damage a white swing does has nothing to do with how much rage it gives - within each group the damage can vary by 2 to 4 times while the rage stays exactly the same.
2. **Crits don't give extra rage.** A crit gives the same rage as a normal hit, and so does a glancing blow.
3. **Rage scales with weapon speed.** The time between swings in each group lines up with one of the posted weapon speeds, and the rage per swing is just that speed times a fixed rate:
   - main hand (one-handed): about **3.46 rage per second of speed**
   - off hand: about **1.73**, which is exactly half the main hand
   - two-handed: about **4.5** (confirmed with two different two-handers, at 3.3 and 3.8 speed)
4. **Weapon speed doesn't change your rage per second.** Since a slow weapon gives more rage per swing but swings less often, fast and slow weapons end up generating rage at the same rate over time.
5. **Level didn't make a visible difference either.** Every character in our logs gets the same rates, from the level 13 ones up to a level 20 one, even though the Classic formula would have changed quite a bit over that range.

### How strong is the evidence? p-values

A p-value answers one question: if an idea were true, how surprising would our data be? Every test starts from an idea (statisticians call it the null hypothesis), and the p-value is the chance of seeing data at least as far from that idea as ours, assuming the idea is actually true.

- A small p-value (below 0.05, and here usually way below) means our data would be really surprising if the idea were true, so we reject it.
- A large p-value means the data fit the idea just fine. That doesn't prove the idea is right - it only means the logs don't give us any reason to doubt it.

Wherever we can, we test each finding twice: once against what Classic predicts, and once against what we think Forever does. One small caveat is that the log rounds rage to one decimal place, so the values aren't a perfect bell curve, but with more than 100 swings in each group that doesn't change the results of these tests.

```{r}
by_group <- obs %>% group_by(group) %>% filter(n() >= 10)

# 1. Damage: slope of rage on damage, against Classic's slope (15 / C) and against zero.
dmg <- by_group %>% group_modify(~ {
  f  <- summary(lm(gain ~ damage, data = .x))$coefficients
  b  <- f["damage", "Estimate"]; se <- f["damage", "Std. Error"]; df <- nrow(.x) - 2
  cs <- mean(15 / conv(.x$level))
  tibble(slope = b, classic_slope = cs,
         p_classic = 2 * pt(-abs((b - cs) / se), df), p_zero = 2 * pt(-abs(b / se), df))
})

# 2. Crits: rage of a crit relative to the group's normal rage, against Classic (2x) and against no change (1x).
crit <- by_group %>% mutate(rel = gain / mean(gain[!crit])) %>% ungroup()
crit_rel <- crit$rel[crit$crit]
p_crit_classic <- t.test(crit_rel, mu = 2)$p.value
p_crit_same    <- t.test(crit$rel ~ crit$crit)$p.value

# 3. Off hand is half the main hand: off-hand rate minus half the main-hand rate, per log.
rates <- obs %>% filter(hand != "Two-hand") %>% mutate(rate = gain / speed)
half <- rates %>% group_by(character, file) %>% filter(n_distinct(hand) == 2) %>%
  group_modify(~ {
    mh <- .x$rate[.x$hand == "Main hand"]; oh <- .x$rate[.x$hand == "Off hand"]
    z  <- (mean(oh) - 0.5 * mean(mh)) / sqrt(var(oh) / length(oh) + 0.25 * var(mh) / length(mh))
    tibble(ratio = mean(oh) / mean(mh), p = 2 * pnorm(-abs(z)))
  })

# 4. Proportional to speed: same rate at different speeds, within one hand.
p_speed_mh <- t.test(rate ~ speed, data = filter(rates, hand == "Main hand", file == meta[[2]]$file))$p.value
p_speed_oh <- oneway.test(rate ~ factor(speed), data = filter(rates, hand == "Off hand"))$p.value

# 5. Level: the same 1.8-speed main hand at level 15 (Siax) and level 16 (Zomd).
lv <- filter(obs, hand == "Main hand", speed == 1.8)
p_level <- t.test(gain ~ level, data = lv)$p.value
classic_gap <- 1 - conv(15) / conv(16)

tribble(
  ~Finding, ~`Idea tested (null hypothesis)`, ~`p-value`, ~Verdict,
  "1. Damage does not change rage", "Classic: rage rises by 15 / C for each point of damage",
    paste("largest of", nrow(dmg), "groups:", fmt_p(max(dmg$p_classic))), "Classic rejected",
  "1. Damage does not change rage", "Forever: damage has no effect on rage",
    paste("smallest of", nrow(dmg), "groups:", fmt_p(min(dmg$p_zero))), "See note below",
  "2. Crits give no extra rage", "Classic: a crit gives twice the rage", fmt_p(p_crit_classic), "Classic rejected",
  "2. Crits give no extra rage", "Forever: a crit gives the same rage", fmt_p(p_crit_same),
    if (p_crit_same > 0.05) "Consistent" else "Rejected",
  "3. Off hand is half the main hand", "Off-hand rate = 0.5 x main-hand rate",
    paste(paste0(half$character, ": ", fmt_p(half$p)), collapse = "; "),
    if (all(half$p > 0.05)) "Consistent" else "See note below",
  "3. Rage is proportional to speed", "Main hand: same rate at 1.8 and 2.6 speed", fmt_p(p_speed_mh),
    if (p_speed_mh > 0.05) "Consistent" else "See note below",
  "3. Rage is proportional to speed", "Off hand: same rate at 1.3, 1.8 and 2.6 speed", fmt_p(p_speed_oh),
    if (p_speed_oh > 0.05) "Consistent" else "See note below",
  "5. Level has no effect", "Level 15 and level 16 give the same rage (1.8 main hand)", fmt_p(p_level),
    if (p_level > 0.05) "Consistent" else "See note below"
) %>% knitr::kable(caption = "p-values for each finding. 'Rejected' means the data are very unlikely if the idea were true. 'Consistent' means the data give no reason to doubt it.")
```

A couple of notes on reading the table:

- **Damage.** Against Classic, the p-values are so small that R just prints them as below $10^{-10}$. Against "no effect" it's a bit trickier, because with lots of swings even a tiny effect can come out as "significant", so it's worth looking at how big the effect actually is. The largest slope we measured is `r round(max(abs(dmg$slope)) * 100, 2)` rage per 100 damage, while Classic predicts about `r round(mean(dmg$classic_slope) * 100, 0)` rage per 100 damage at these levels - more than `r floor(min(dmg$classic_slope / pmax(abs(dmg$slope), 1e-9)))` times larger. So if damage does have any effect at all, it's far too small to matter.
- **Level.** Classic predicts that the same damage gives `r round(classic_gap * 100, 1)`% less rage at level 16 than at level 15, and the logs show a difference of just `r round(abs(diff(tapply(lv$gain, lv$level, mean))) / mean(lv$gain) * 100, 2)`%. That said, 13 to 16 is a pretty small range (and levels 15 and 16 are also two different characters), so we'll need logs from higher levels to be confident about this one.

#### What level does to rage

To look at the effect of level on its own, we need to keep everything else the same. So we'll take one weapon (Tail Spike, 1.8 speed, in the main hand) and one hit of `r round(median(lv$damage))` damage, which is the typical logged hit for this weapon, and only change the level.

```{r}
d_ref <- median(lv$damage)
lv_obs <- lv %>% group_by(level) %>%
  summarise(mean = mean(gain), lo = mean - 1.96 * sd(gain) / sqrt(n()), hi = mean + 1.96 * sd(gain) / sqrt(n()),
            n = n(), .groups = "drop") %>%
  mutate(game = "Forever (logs)")
lv_classic <- tibble(level = c(15, 16), mean = 15 * d_ref / conv(level), game = "Classic (formula)")
```

```{r fig.cap="Levels 15 and 16, with the same weapon and the same hit. Classic (grey) drops by about 6% in a single level, while the logs (blue, with 95% intervals too narrow to see) don't move.", fig.height=3.5}
ggplot(mapping = aes(factor(level), mean, colour = game, group = game)) +
  geom_line(data = lv_classic, linewidth = 1) + geom_point(data = lv_classic, size = 3) +
  geom_line(data = lv_obs, linewidth = 1) +
  geom_pointrange(data = lv_obs, aes(ymin = lo, ymax = hi), size = 0.6) +
  geom_text(data = lv_obs, aes(label = paste(n, "swings")), vjust = -1.2, show.legend = FALSE) +
  scale_colour_manual(values = c("Classic (formula)" = "grey50", "Forever (logs)" = "#2c7fb8")) +
  labs(x = "Level", y = "Rage for one swing", colour = NULL)
```

The two levels we have logs for barely show anything, so here's the same comparison stretched out from level 10 to 60. Keep in mind that the damage is frozen at that same `r round(median(lv$damage))`-damage hit the whole way - the plot isn't asking how much rage a warrior gets at each level, but how much rage *this particular hit* is worth at each level. The solid part of the Forever line is where we actually have logs, and the dashed part is us assuming it stays flat.

```{r fig.cap="How much rage the same 1.8-speed main-hand hit is worth at each level. In Classic (grey) it's worth less and less as you level, because C keeps growing, while in Forever (blue) it stays the same. The dots are the logs, and the circle is where both games value the hit the same.", fig.height=4}
curve <- tibble(level = seq(10, 60, by = 0.25)) %>%
  mutate(Classic = 15 * d_ref / conv(level),
         Forever = mean(lv$gain))
seen <- range(obs$level)
cross <- curve$level[which.min(abs(curve$Classic - curve$Forever))]

ggplot(curve, aes(level)) +
  annotate("rect", xmin = seen[1], xmax = seen[2], ymin = -Inf, ymax = Inf, fill = "#2c7fb8", alpha = 0.08) +
  annotate("text", x = mean(seen), y = max(curve$Classic) * 0.95, label = "logged\nlevels", size = 3, colour = "#2c7fb8") +
  geom_line(aes(y = Classic, colour = "Classic (formula)"), linewidth = 1) +
  geom_line(data = filter(curve, level >= seen[1], level <= seen[2]), aes(y = Forever, colour = "Forever"), linewidth = 1.2) +
  geom_line(aes(y = Forever, colour = "Forever"), linetype = "dashed") +
  geom_point(data = lv_obs, aes(level, mean, colour = "Forever"), size = 2.5) +
  scale_colour_manual(values = c("Classic (formula)" = "grey50", "Forever" = "#2c7fb8")) +
  annotate("point", x = cross, y = mean(lv$gain), shape = 21, size = 4, stroke = 1) +
  annotate("text", x = cross, y = mean(lv$gain), label = paste("same rage at level", round(cross)),
           vjust = -1.5, size = 3) +
  scale_x_continuous(breaks = seq(10, 60, 10)) +
  labs(x = "Level", y = "Rage for one swing", colour = NULL)
```

In Classic, that hit is worth `r round(15 * d_ref / conv(15), 1)` rage at level 15 but only `r round(15 * d_ref / conv(60), 1)` at level 60 (`r round((1 - conv(15) / conv(60)) * 100)`% less), and the two games value it the same at around level `r round(cross)` - below that Classic gives more for this hit, and above it Forever does. This doesn't mean Classic warriors were starved for rage at 60, though: by then your hits do far more than `r round(median(lv$damage))` damage, and that extra damage is what makes up for the shrinking conversion. In Forever, on the other hand, damage doesn't count at all, so hitting harder doesn't get you any more rage.

## Classic versus Forever

We have no Classic logs in this format to compare against (yet). Instead, we can ask what the Classic formula would have given for the exact same swings: same damage, same character level.

```{r fig.cap="Rage per swing under the Classic formula (grey) against what the logs show (colour), for the same swings. Classic spreads with damage; Forever is a spike per weapon."}
obs %>%
  select(hand, speed, Classic = classic, Observed = gain) %>%
  pivot_longer(c(Classic, Observed), names_to = "source", values_to = "rage") %>%
  ggplot(aes(rage, fill = source)) +
  geom_histogram(binwidth = 0.5, position = "identity", alpha = 0.7) +
  facet_wrap(~ paste0(hand, " (", speed, " speed)"), scales = "free_y") +
  scale_fill_manual(values = c(Classic = "grey60", Observed = "#2c7fb8")) +
  labs(x = "Rage per swing", y = "Swings", fill = NULL)
```

For these exact hits, the Classic formula would have given `r round(mean(obs$classic), 1)` rage per swing on average, against the `r round(mean(obs$gain), 1)` we actually saw. The shape matters more than the average, though - under Classic rage follows damage and a crit roughly doubles it, while in Forever **rage per swing does not move**.

### How sure are we? A Bayesian estimate

The findings above give one number for each hand. This section asks a second question: how sure are we of each number? We can use Bayesian statistics to answer it. Bayesian methods starts from what we **believe** before we look at the data, and then updates that belief with the data once we have it.

The model is the equation that we think explains the rage of each swing:

\begin{equation}
Rage = r_{hand} \times speed + \varepsilon
\end{equation}

The terms have these meanings:

- $Rage$ is the rage that one white swing gives.
- $speed$ is the speed of the weapon that made the swing, as the tooltip shows it. For example, 2.6 for Vileblood Scimitar.
- $r_{hand}$ is the rate: the rage for each second of weapon speed. It has one value for the main hand, one for the off hand, and one for two-handed weapons. We want to find these three values.
- $\varepsilon$ (epsilon) is the noise. It is the small difference between the model and each real swing. The game shows rage to one decimal place, so each swing is a little above or below the model.

There are three important terms that we need to become familiar with when we talk about Bayesian bayesian statistics:

- The **prior** is what we believe about the rate before we look at the logs. We used a flat prior, which just gives every value between 1 and 6 the same chance. It tells the model that we have no preferred answer, so the result comes from the logs alone.
- The **posterior** is what we believe after we look at the logs. It gives each possible rate a probability, so a number from 0 to 1 that shows how likely a value is.
- The **95% interval** is the range that holds the true rate with a probability of 95%.

To find the posterior, the computer tries every rate from 1 to 6 in small steps. For each rate, it calculates how well that rate explains all the swings in the logs. Rates that explain the swings well get a high probability. This method is called grid approximation.

We then use the posterior to simulate new swings at weapon speeds that nobody has logged yet. If the simulated swings agree with the real ones, the model is good: this check is what we call the posterior predictive.

```{r}
grid <- seq(1, 6, by = 0.0005)
post <- obs %>% group_by(hand) %>%
  group_modify(~ {
    sig <- sd(.x$gain - (.x$gain / .x$speed) %>% mean() * .x$speed) %>% max(0.05)
    ll  <- map_dbl(grid, function(r) sum(dnorm(.x$gain, r * .x$speed, sig, log = TRUE)))
    p   <- exp(ll - max(ll)); p <- p / sum(p)
    tibble(rate = sample(grid, 4000, replace = TRUE, prob = p), sigma = sig)
  })

```

Looking at the logs, the formula comes down to nice round numbers: every white swing gives **3.5 × weapon speed** rage in the main hand, **half of that (1.75 × speed)** in the off hand, and **4.5 × speed** with a two-hander. (If you want to be picky, the one-handed rates actually come in about 1% lower, at 3.46 and 1.73, but the round numbers are the easiest way to remember it.)

```{r fig.cap="The posterior predictive. Each line shows the rage that the model predicts for one swing at each weapon speed. The shaded band shows where 95% of new swings will fall. The dots are the real swings from the logs."}
sp <- seq(1.2, 4, by = 0.1)
pp <- post %>% group_by(hand) %>% slice_sample(n = 500) %>%
  crossing(speed = sp) %>%
  filter(if_else(hand == "Two-hand", speed >= 2.6, speed <= 3)) %>%
  mutate(sim = rnorm(n(), rate * speed, sigma)) %>%
  group_by(hand, speed) %>%
  summarise(lo = quantile(sim, 0.025), mid = median(sim), hi = quantile(sim, 0.975), .groups = "drop")

ggplot(pp, aes(speed, mid, colour = hand, fill = hand)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line() +
  geom_jitter(data = obs, aes(speed, gain), width = 0.02, height = 0, alpha = 0.3, size = 0.8) +
  labs(x = "Weapon speed", y = "Rage per swing", colour = NULL, fill = NULL)
```

The intervals are very narrow (basically impossible to even see), because the rage from each weapon almost never changes, meaning we are pretty sure of each rate. What we used to be less sure of was the model itself - all two-handed swings came from one weapon, with a speed of 3.8 (one point does not show the shape of a line, obviously), so two-handed weapons could have had a rate of 4.5, or the one-handed rate plus a fixed bonus for each swing. Tarch's Rockslicer (3.3 speed) now gives us a second point, and it lands right on 4.5 × speed: with a flat bonus instead, a 3.3-speed two-hander would give about `r round(mean(post$rate[post$hand == "Main hand"]) * 3.3 + (mean(post$rate[post$hand == "Two-hand"]) - mean(post$rate[post$hand == "Main hand"])) * 3.8, 1)` rage per swing, not the `r round(mean(obs$gain[obs$hand == "Two-hand" & obs$speed == 3.3]), 2)` we see.

## Level 60: Classic versus Forever

Our logs come from characters between level 13 and 16, but what most people actually care about is level 60. So the question for this section is: with the same weapons, how much rage would a level 60 warrior get in Classic, and how much in Forever?

We can't know the exact gear of "a level 60 warrior", so instead of picking one number for each unknown we give each one a prior (a range of likely values). The computer then plays 4000 level 60 warriors, each with values drawn at random from those priors, which gives us a whole distribution of rage rather than a single number - and the spread of that distribution shows how uncertain we are.

These are the priors we used:

- **Attack power**: around 1200, usually somewhere between 900 and 1500, which is roughly a fury warrior in early raid gear.
- **Crit chance**: around 30% (the same number the Classic compendium used), usually between 21% and 39%.
- **Boss armor**: a boss with Sunder Armor and Faerie Fire up takes about 75% of full damage, so we use a range of roughly 65% to 83%.
- **Weapon damage**: taken from the logs. Each simulated swing starts from a real hit with that weapon in that hand, and then we add the damage from the extra attack power you'd have at 60 (in Classic, that's 1 damage for every 14 attack power, for each second of weapon speed).
- **The Forever rates**: drawn from the posterior in the section above.

Classic then uses its formula with the level 60 conversion value ($C = `r round(conv(60), 1)`$), while Forever just uses rate × speed. We're only counting hits that land here, so misses, dodges and glancing blows aren't included. The logged hits were also against low-level mobs that already have a bit of armor, and we put the boss armor on top of that, so the Classic side ends up slightly lower than it really would be.

There's one big assumption in all of this: our Forever rates come from levels 13 to 16, and we're assuming they don't change all the way up to 60. The logs can't test that yet, so if the rates do turn out to change with level, the Forever side of these plots will move.

```{r}
n_sim <- 4000
C60 <- conv(60)
setups <- tribble(
  ~setup,                  ~hand,       ~weapon,              ~speed,
  "Siax: Tail Spike / Poniard",     "Main hand", "Tail Spike",         1.8,
  "Siax: Tail Spike / Poniard",     "Off hand",  "Poniard",            1.3,
  "Zomd: Vileblood Scimitar / Tail Spike", "Main hand", "Vileblood Scimitar", 2.6,
  "Zomd: Vileblood Scimitar / Tail Spike", "Off hand",  "Tail Spike",         1.8,
  "Zomd: Battle Axe (2H)",         "Two-hand",  "Battle Axe",         3.8
)

# Real logged non-crit hits for each weapon in each hand, with the attack power at the time.
logged <- obs %>% filter(!crit, !glancing) %>% select(weapon, hand, damage, ap)
pick_hit <- function(w, h) { x <- logged[logged$weapon == w & logged$hand == h, ]; x[sample(nrow(x), 1), ] }

draws <- tibble(sim = 1:n_sim,
                ap    = rnorm(n_sim, 1200, 150),
                crit  = rbeta(n_sim, 30, 70),
                armor = rbeta(n_sim, 75, 25))

forever_rate <- post %>% group_by(hand) %>% slice_sample(n = n_sim, replace = TRUE) %>%
  mutate(sim = row_number()) %>% ungroup() %>% select(hand, sim, rate, sigma)

sim <- setups %>% crossing(draws) %>%
  left_join(forever_rate, by = c("hand", "sim")) %>%
  mutate(hm = if_else(hand == "Off hand", 0.5, 1),
         hit = map2(weapon, hand, pick_hit),
         log_damage = map_dbl(hit, "damage"), log_ap = map_dbl(hit, "ap"),
         # Classic: 1 extra damage for every 14 attack power, for each second of weapon speed.
         hit_damage = pmax(log_damage + hm * (ap - log_ap) / 14 * speed, 1) * armor,
         is_crit = runif(n()) < crit,
         damage = hit_damage * if_else(is_crit, 2, 1),
         Classic = 15 * damage / C60,
         Forever = rate * speed + rnorm(n(), 0, sigma))
```

```{r}
# Forever is almost a single value, so it is drawn as a line with its 95% interval as a band.
# Classic is drawn as a distribution scaled to the same height, so both are easy to see.
compare_plot <- function(d, facet, x_lab) {
  fv <- d %>% group_by({{ facet }}) %>%
    summarise(mid = median(Forever), lo = quantile(Forever, .025), hi = quantile(Forever, .975))
  ggplot(d, aes(Classic)) +
    geom_density(aes(y = after_stat(scaled), fill = "Classic"), colour = NA, alpha = 0.6) +
    geom_rect(data = fv, aes(xmin = lo, xmax = hi, ymin = 0, ymax = 1, fill = "Forever"),
              inherit.aes = FALSE, alpha = 0.6) +
    geom_vline(data = fv, aes(xintercept = mid), colour = "#2c7fb8", linewidth = 1) +
    facet_wrap(vars({{ facet }}), ncol = 1) +
    scale_fill_manual(values = c(Classic = "grey60", Forever = "#2c7fb8")) +
    labs(x = x_lab, y = NULL, fill = NULL) +
    theme(axis.text.y = element_blank(), panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank())
}
```

```{r fig.cap="Rage for one landed swing at level 60, for each logged weapon. Grey: Classic. It spreads out because rage follows damage, and crits form the second bump on the right. Blue line: Forever. It is the same for every swing.", fig.height=7}
sim %>% mutate(label = paste0(weapon, " (", hand, ", ", speed, ")")) %>%
  compare_plot(label, "Rage for one landed swing")
```

Rage per swing is a bit misleading when comparing weapons, since slow weapons swing less often. A better way to compare them is rage per second - all the rage from your swings over one second of combat, adding up both hands if you're dual wielding.

```{r}
per_sec <- sim %>%
  mutate(crit_mult = 1 + crit,  # the average of 1x and 2x damage, at this crit chance
         classic_ps = 15 * hit_damage * crit_mult / C60 / speed,
         forever_ps = rate) %>%
  group_by(setup, sim) %>%
  summarise(Classic = sum(classic_ps), Forever = sum(forever_ps), .groups = "drop")
```

```{r fig.cap="Rage per second at level 60 from white hits, for each weapon setup. Grey: Classic, where each simulated warrior has different gear. Blue line: Forever. It does not change with gear or weapon speed, so it is in the same place for both dual-wield setups."}
compare_plot(per_sec, setup, "Rage per second from white hits")
```

```{r}
per_sec %>% group_by(Setup = setup) %>%
  summarise(`Classic (median)` = median(Classic),
            `Classic (95% interval)` = sprintf("%.1f to %.1f", quantile(Classic, .025), quantile(Classic, .975)),
            `Forever (median)` = median(Forever),
            `Probability that Forever gives less` = mean(Forever < Classic)) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  knitr::kable(caption = "Rage per second from white hits at level 60. The last column is the share of simulated warriors who get less rage in Forever than in Classic.")
```

So what actually changes for a level 60 warrior?

- In Classic, better gear means more rage, since more attack power and more crits mean more damage. In Forever, your gear doesn't change the rage you get from white hits at all.
- In Classic, your rage per second changes from fight to fight and from player to player, whereas in Forever it's basically fixed: about `r round(mean(post$rate[post$hand == "Main hand"]) + mean(post$rate[post$hand == "Off hand"]), 2)` for dual wield and `r round(mean(post$rate[post$hand == "Two-hand"]), 2)` for a two-hander.
- The weapons in these logs are low-level ones, and real level 60 weapons hit a lot harder. That extra damage means more rage in Classic but not in Forever, so with actual level 60 weapons the gap would be even bigger than what these plots show.

## Other rage sources

```{r}
energize %>% filter(amount > 0) %>% mutate(Log = str_remove(file, "_092326.txt")) %>% group_by(Log, spell) %>%
  summarise(`Typical amount` = amount[which.max(n)], Events = sum(n)) %>%
  knitr::kable(caption = "Rage granted directly by abilities and talents (amounts below the typical value were cut off by the cap).")
```

- **Charge** gives 9 rage and **Bloodrage** gives 10 up front plus 1 per tick, same as in Classic.
- **Unbridled Wrath** gives 1 rage per proc with one-handers but 2 with the two-hander, which fits the same pattern of two-handers getting more rage per event.
- **Unbridled Wrath procs a lot less often than its tooltip says** (a bug), and never off Heroic Strike or Cleave. There's a whole section on it below.
- **Heroic Strike** hits don't give any rage, at least in the cases we checked.
- **Misses, dodges and parries** on our own swings give little or no rage (though we only have a few of them).
- **Rage from damage taken** is small, but it varied between logs, so it needs a proper controlled test.

### Unbridled Wrath

Unbridled Wrath is meant to give a 12% chance per talent point to gain rage on a white hit, so 60% at 5/5. To test that, Tarch recorded the same fights with 1, 2, 3 and 4 points in the talent (and nothing else but 5/5 Cruelty), and Zomd's logs were recorded at 5/5. That gives us the proc rate at every talent rank:

```{r}
ubw <- procs %>% filter(!is.na(ubw_rank)) %>%
  group_by(rank = ubw_rank) %>%
  summarise(n = sum(landed_white), k = sum(ubw_procs), .groups = "drop")

# Best estimate of the chance per talent point, assuming each point adds the same chance
# (maximum likelihood over all ranks together).
per_point <- optimize(function(q) -sum(dbinom(ubw$k, ubw$n, pmin(q * ubw$rank, 1), log = TRUE)),
                      c(0.01, 0.2))$minimum

ubw %>%
  transmute(`Talent points` = paste0(rank, "/5"), `Landed white hits` = n, Procs = k,
            `Proc rate` = sprintf("%.1f%%", k / n * 100),
            `95% interval` = sprintf("%.1f%% to %.1f%%", qbeta(0.025, k + 1, n - k + 1) * 100,
                                     qbeta(0.975, k + 1, n - k + 1) * 100),
            `Expected at 12% per point` = sprintf("%.0f%%", 12 * rank),
            `p-value vs 12% per point` = map2_chr(k, n, ~ fmt_p(binom.test(.x, .y, 0.12 * ubw$rank[ubw$k == .x & ubw$n == .y])$p.value)),
            `Expected at 8% per point` = sprintf("%.0f%%", 8 * rank),
            `p-value vs 8% per point` = map2_chr(k, n, ~ fmt_p(binom.test(.x, .y, 0.08 * ubw$rank[ubw$k == .x & ubw$n == .y])$p.value))) %>%
  knitr::kable(caption = "How often Unbridled Wrath procs on a landed white hit, by talent points. The 95% interval comes from a Bayesian posterior with a flat prior. A small p-value means the logs are very unlikely at that proc rate.")
```

```{r fig.cap="Unbridled Wrath's proc rate at each talent rank (dots, with 95% intervals), against 12% per point (what the talent should give, dashed) and 8% per point (dotted). The logs follow the lower line.", fig.height=3.5}
ubw %>% mutate(rate = k / n, lo = qbeta(0.025, k + 1, n - k + 1), hi = qbeta(0.975, k + 1, n - k + 1)) %>%
  ggplot(aes(rank, rate)) +
  geom_abline(slope = 0.12, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_abline(slope = 0.08, intercept = 0, linetype = "dotted", colour = "grey50") +
  annotate("text", x = 4.6, y = 0.12 * 4.6, label = "12% per point", hjust = 1, vjust = -0.6, size = 3, colour = "grey40") +
  annotate("text", x = 4.6, y = 0.08 * 4.6, label = "8% per point", hjust = 0, vjust = 1.6, size = 3, colour = "grey40") +
  geom_pointrange(aes(ymin = lo, ymax = hi), colour = "#2c7fb8") +
  scale_x_continuous(breaks = 1:5, labels = paste0(1:5, "/5"), limits = c(0.8, 5.2)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.65)) +
  labs(x = "Points in Unbridled Wrath", y = "Procs per landed white hit")
```

The talent clearly doesn't give 12% per point: that's rejected at every rank (only just at 1/5, where we have the fewest procs). Our best estimate is about **`r round(per_point * 100, 1)`% per point**, and 5/5 lands at `r round(ubw$k[ubw$rank == 5] / ubw$n[ubw$rank == 5] * 100)`%, a long way from 60%. The tooltip says 12% per point, so this is a bug: it's close to the 8% per point Classic's Unbridled Wrath had (40% at 5/5), so the beta may still be using the old Classic chance. It's listed as [BUG-1](#bug-ubw) in the Known Bugs chapter. Siax's rank wasn't posted, but his proc rate (`r round(procs$ubw_procs[procs$character == "Siax"] / procs$landed_white[procs$character == "Siax"] * 100)`%) matches 5/5.

Two more things stay the same as before: every proc gives 1 rage with one-handers and 2 with two-handers, at every rank, and none of the procs come from Heroic Strike or Cleave.

## Open questions

- Check whether the rates change at higher levels.
- Figure out how attack power adds damage in Forever. The logs show 380 to 480 attack power at levels 13 to 16, and with the Classic rule (1 damage for every 14 attack power, for each second of weapon speed) that should add about 100 damage to every Battle Axe swing - but the logged Battle Axe hits are only 69 to 88. So either attack power works differently, or the number in the log means something else.
- Keep checking new builds to see when [BUG-1](#bug-ubw) gets fixed.
- Test Dual Wield Specialization, which Bored's spreadsheet assumes doubles off-hand rage, with a log posted with talents.
- Test stances, and rage from damage taken, in a controlled setting.
- Record mid-log hand swaps in the metadata, since one log shows both weapons being used in both hands.

## Changelog {#rage-changelog}

This chapter will keep changing as more logs come in, so each version of the findings gets a number in the form MAJOR.MINOR.PATCH, plus the date it was published. The changelog only tracks the findings, so rewording or tidying up the text doesn't get a new version.

- **MAJOR** goes up when a conclusion changes, for example if a new log shows that rage does depend on damage after all, or that a rate is different from what we thought.
- **MINOR** goes up when we add findings without overturning what was there, like new logs that agree with the current results, or a new mechanic we've measured.
- **PATCH** goes up when we correct a number (a better estimate, or a fixed calculation) without changing any conclusion.

```{r}
changelog %>% rename(Version = version, Date = date, `What changed` = change) %>%
  knitr::kable(caption = "Versions of this chapter, newest first.")
```
