# Known Bugs {#known-bugs}

```{r setup-bugs, include=FALSE}
# Reuses `procs`, `ubw`, `per_point`, `ubw_rate` and `fmt_p` from the Rage Generation chapter.
source("data/fight_sim.R")  # for the weapon speeds and attack table used in the cost estimate
bugs <- tribble(
  ~id,     ~bug,                                                    ~build,   ~first_seen,  ~status,
  "BUG-1", "Unbridled Wrath procs about 8% per point instead of 12%", "1.60.1", "2026-09-23", "Open"
)
```

This chapter keeps track of everything we've found where the beta doesn't do what the game says it should - usually a tooltip promising one number while the logs show another. We only list something here when the logs make the case clearly (a p-value well below 0.05 against the tooltip value, over a decent number of events), and each bug gets an ID, the build it was seen on and a status, so we can tell when it gets fixed.

```{r}
bugs %>% rename(ID = id, Bug = bug, Build = build, `First seen` = first_seen, Status = status) %>%
  knitr::kable(caption = "Known bugs, newest first.")
```

If you've found something that looks wrong, the best way to report it is to post a log with the talents and gear involved (see the [log guidelines](https://github.com/magey/forever-warrior/discussions)), so it can be checked the same way.

## BUG-1: Unbridled Wrath procs too rarely {#bug-ubw}

**What the tooltip says:** Unbridled Wrath gives a 12% chance per talent point to gain rage on a white hit, so 60% at 5/5.

**What the logs show:** about `r round(per_point * 100, 1)`% per point, so 5/5 only procs on about `r round(ubw$k[ubw$rank == 5] / ubw$n[ubw$rank == 5] * 100)`% of landed white hits. That's roughly Classic's old value of 8% per point, so it looks like the beta is still using the Classic proc chance while the tooltip shows the new one.

```{r}
ubw %>%
  transmute(`Talent points` = paste0(rank, "/5"),
            `Tooltip chance` = sprintf("%.0f%%", 12 * rank),
            `Measured chance` = sprintf("%.1f%%", k / n * 100),
            `95% interval` = sprintf("%.1f%% to %.1f%%", qbeta(0.025, k + 1, n - k + 1) * 100, qbeta(0.975, k + 1, n - k + 1) * 100),
            `Landed white hits` = n,
            `p-value vs tooltip` = map2_chr(k, n, ~ fmt_p(binom.test(.x, .y, 0.12 * ubw$rank[ubw$k == .x & ubw$n == .y])$p.value))) %>%
  knitr::kable(caption = "Unbridled Wrath's proc chance: tooltip against logs. The full breakdown is in the Rage Generation chapter.")
```

**How sure are we:** very. The tooltip value is rejected at every rank we've logged (only just at 1/5, where there are the fewest procs, and overwhelmingly at 5/5), and the procs follow roughly 8% per point from 1/5 to 3/5. The top two ranks come in close together (4/5 has fewer events, so its range is wide), with 5/5 a little under the 40% that 8% per point would give. The rage per proc hasn't changed between logs: 1 rage with one-handers and 2 with two-handers, at every rank.

**What it costs you:** rage from Unbridled Wrath scales with how many white hits you land, so the bug costs about `r round((0.60 - ubw_rate) * 60, 1)` procs for every 60 landed white hits. For a level 60 dual wielder with the weapons from the fight simulation (a `r sim_defaults$mh_speed` and a `r sim_defaults$oh_speed` speed weapon, landing about `r round((1 - sim_defaults$miss - sim_defaults$dodge) * 100)`% of swings), that's roughly **`r round((0.60 - ubw_rate) * (1 - sim_defaults$miss - sim_defaults$dodge) * (1 / sim_defaults$mh_speed + 1 / sim_defaults$oh_speed) * 60, 1)` rage per minute** lost, and for a 3.3-speed two-hander (2 rage per proc) about `r round((0.60 - ubw_rate) * (1 - sim_defaults$miss - sim_defaults$dodge) * (1 / 3.3) * 2 * 60, 1)` rage per minute. Not huge, but for a class that's already short on rage, it adds up.

**Seen on:** build 1.60.1, in logs from Siax, Zomd Aya and Tarch (Unbridled Wrath at every rank from 1/5 to 5/5).

**Status:** open. The proposals chapter uses the tooltip's 60%, since that's how the game is meant to work.
